A Practical Optional Type System for Clojure

### 2012-11-26 dissertation/design-choices.tex
Correct typos in dissertation
Chris' corrections

This chapter describes the design choices made for Typed Clojure.
Typed Clojure is designed to be of practical use to Clojure programmers.
To this end, many of the design choices are borrowed from other existing projects.
Some of the main design choices that specifically concern Typed Clojure are given more
more detailed descriptions.

@@

\begin{lstlisting}[caption=Type assiciated with clojure.core/seq, label=lst:seqtype]
(All [x]
  [(U nil (Seqable x)) -> (U nil (Seqable x))
   :filters {:then (is (CountRange 1) 0)
             :else (| (is nil 0)
                      (is (ExactCount 0) 0))}])
\end{lstlisting}

@@

Correspondingly, \lstinline|first| which gets the first member of a sequence.
can be given a more accurate (and surprisingly subtle) type.
\lstinline|first| is overloaded in a similar fashion to \lstinline|seq|.
\lstinline|first| takes an argument that is either a sequence or \lstinline|nil|.
If it is passed \lstinline|nil| or an empty sequence, it returns \lstinline|nil|. Otherwise, if it is passed
non-empty sequence, it returns the first member of the sequence.
Listing \ref{lst:firsttype} shows the type of \lstinline|first|.

\begin{lstlisting}[caption=Type assiciated with clojure.core/first, label=lst:firsttype]
(All [x]
  (Fn [(U nil (I (Seqable x) (ExactCount 0))) -> nil]
      [(I (Seqable x) (CountRange 1)) -> x]
      [(U nil (Seqable x)) -> (U nil x)])))
\end{lstlisting}

@@

Primitive arrays in Java are covariant, which is well-known to be statically unsound~\cite{OW97,BK99}.
This mean that at compile time we cannot completely trust the type signature of any Java method or field that involves 
an array.

@@

Protocols and datatypes are two of Clojure's main means of abstraction.
Protocols, have a set of unimplemented methods, are typed by
giving the type for each method.
Datatype definitions are type checked by ensuring the methods it
implements match the types defined by the Java interface or protocol
it is extending.

>>>>

This chapter describes the design choices made for Typed Clojure.
Typed Clojure is designed to be of practical use to Clojure programmers.
To this end, many of the design choices are borrowed from other existing projects.
Some of the main design choices that specifically concern Typed Clojure are given
more detailed descriptions.

@@

\begin{lstlisting}[caption=Type associated with clojure.core/seq, label=lst:seqtype]
(All [x]
  [(U nil (Seqable x)) -> (U nil (Seqable x))
   :filters {:then (is (CountRange 1) 0)
             :else (| (is nil 0)
                      (is (ExactCount 0) 0))}])
\end{lstlisting}

@@

Correspondingly, \lstinline|first| which returns the first member of a sequence,
can be given a more accurate (and surprisingly subtle) type.
\lstinline|first| is overloaded in a similar fashion to \lstinline|seq|.
\lstinline|first| takes an argument that is either a sequence or \lstinline|nil|.
If it is passed \lstinline|nil| or an empty sequence, it returns \lstinline|nil|. Otherwise, if it is passed
non-empty sequence, it returns the first member of the sequence.
Listing \ref{lst:firsttype} shows the type of \lstinline|first|.

\begin{lstlisting}[caption=Type associated with clojure.core/first, label=lst:firsttype]
(All [x]
  (Fn [(U nil (I (Seqable x) (ExactCount 0))) -> nil]
      [(I (Seqable x) (CountRange 1)) -> x]
      [(U nil (Seqable x)) -> (U nil x)])))
\end{lstlisting}

@@

Primitive arrays in Java are covariant, which is well-known to be statically unsound~\cite{OW97,BK99}.
This means that at compile time we cannot completely trust the type signature of any Java method or field that involves 
an array.

@@

Protocols and datatypes are two of Clojure's main means of abstraction.
Protocols define a set of unimplemented methods, and each must be assigned types.
Datatype definitions are type checked by ensuring the methods it
implements match the types defined by the Java interface or assigned to the protocol method
it is extending.

### 2012-11-26 dissertation/experiments.tex
Chris' corrections

Listing \ref{lst:monadalias} captures the abstract definition of a monad.
A \lstinline|Monad| is a type constructor parameterised by \lstinline|m|, which is
a type constructor taking a single argument \lstinline|x| (a type) and returning a type 
(written \lstinline|Any|)\footnote{Haskell-like syntax is helpful here, the kind \lstinline|(TFn [[x :variance :covariant]] Any)|
is approximately \lstinline|* -> *|. A more streamlined syntax is planned for future work.}.
The body of the type constructor uses \lstinline|m| abstractly. Typed Clojure
ensures the correct number of arguments are passed and recognises the declared variances
for each parameter, In this case the first argument of \lstinline|m| is declared
a covariant position. When instantiated, \lstinline|m| must also be a type operator
of one covariant argument returning a type.

>>>>

Listing \ref{lst:monadalias} captures the abstract definition of a monad.
A \lstinline|Monad| is a type constructor parameterised by \lstinline|m|, which is
a type constructor taking a single argument \lstinline|x| (a type) and returning a type 
(written \lstinline|Any|)\footnote{Haskell-like syntax is helpful here, the kind \lstinline|(TFn [[x :variance :covariant]] Any)|
is approximately \lstinline|* -> *|. A more streamlined syntax is planned for future work.}.
The body of the type constructor uses \lstinline|m| abstractly. Typed Clojure
ensures the correct number of arguments are passed and recognises the declared variances
for each parameter. In this case the first argument of \lstinline|m| is declared
a covariant position. When instantiated, \lstinline|m| must also be a type operator
of one covariant argument returning a type.

### 2012-11-26 dissertation/intro.tex
Chris' corrections

\begin{lstlisting}[caption=Typed Hello world, label=lst:helloworld]
(ns typed.test.hello-world
  (:require [typed.core :refer [check-ns]]]))

@@

uses \lstinline|typed.core/ann-form| to assign an expected to
the first argument,
which is sufficient to infer the result type

@@

Such annotations are in a sense assumptions; should they turn out to be wrong, Typed Clojure
will infer incorrect types. Typed Racket generates runtime contracts based on static types to ensure
that type errors are caught at runtime, and the original source of the error is blamed~\cite{WF09}.
A similar system for verifying that method annotations like \lstinline|non-nil-return|
actually describe correct types is planned as future work
(see section \ref{sec:contractsblame}).

>>>>

\begin{lstlisting}[caption=Typed Hello world, label=lst:helloworld]
(ns typed.test.hello-world
  (:require [typed.core :refer [check-ns]]))

@@

uses \lstinline|typed.core/ann-form| to assign an expected type to
the first argument,
which is sufficient to infer the result type

@@

Such annotations are in a sense assumptions; should they turn out to be wrong, Typed Clojure
will infer incorrect types. 
This is somewhat troubling, but type safety can be restored by using other techniques.
For example, Typed Racket generates runtime contracts based on static types to ensure
that type errors are caught at runtime, while giving good error messages 
which indicate (or ``blame''~\cite{WF09}) the original source of the error.
A similar system for verifying that method annotations like \lstinline|non-nil-return|
are correct is planned as future work
(see section \ref{sec:contractsblame}).

### 2012-11-26 dissertation/lit-review.tex
Chris' corrections

Rossberg~\cite{Ros07} describes the advantages of having a type
\emph{Dynamic} over other approaches to making statically typed languages
more like dynamic ones and surveys work in this area. 
He says that type \emph{Dynamic}
allows aspects of dynamic style to be represented in statically
typed languages without compromising the type system.
However, he claims type \emph{Dynamic} is rarely used in practical languages
because it is too inconvenient in real programs.
He claims to have fixed these issues, but a practical language has yet to integrated 
his changes.

@@

Gradual typing combines static and runtime type checking so programmers
can the most appropriate one for the situation.
A key feature of gradual type systems is its use of \emph{bidirectional checking}
like~\cite{PT00}, which requires type annotations in natural places like functions.
Type errors in gradual type systems are often more manageable than in Soft Typing and SBA,
often more comparable to type errors in statically typed languages.

@@

SML CIDRE \cite{Dav05} was also a secondary influence - particularly it's use of bidirectional 
checking of extrinsic types including subtyping, intersection types, and it's aim to capture program invariants.  
However, SML CIDRE starts from the intrinsically typed language SML which already has a rich static 
type system, which leads to many different considerations compared to starting from a dynamically typed language.  
Much of this influence was via a student-supervisor relationship, making it particularly hard to pin down the specifics.

>>>>

Rossberg~\cite{Ros07} describes the advantages of having a type
\emph{Dynamic} over other approaches to making statically typed languages
more like dynamic ones and surveys work in this area. 
He says that type \emph{Dynamic}
allows aspects of dynamic style to be represented in statically
typed languages without compromising the type system.
However, he claims type \emph{Dynamic} is rarely used in practical languages
because it is too inconvenient in real programs.
He claims to have fixed these issues, but a practical language has yet to integrate
his changes.

@@

Gradual typing combines static and runtime type checking, so programmers
can utilise whichever is most appropriate for a given situation.
A key feature of gradual type systems is its use of \emph{bidirectional checking}
like~\cite{PT00}, which requires type annotations in natural places like functions.
Type errors in gradual type systems are often more manageable than in Soft Typing and SBA,
often more comparable to type errors in statically typed languages.

@@

SML CIDRE \cite{Dav05} was also a secondary influence - particularly its use of bidirectional 
checking of extrinsic types including subtyping, intersection types, and its aim to capture program invariants.  
However, SML CIDRE starts from the intrinsically typed language SML which already has a rich static 
type system, which leads to many different considerations compared to starting from a dynamically typed language.  
Much of this influence was via a student-supervisor relationship, making it particularly hard to pin down the specifics.

### 2012-11-26 dissertation/paper.tex
Chris' corrections

Typed Clojure is intended to be of practical use to Clojure programmers.
We show that Typed Clojure helps verify the absence of errors in Clojure code written in sophisticated
programming styles by porting most of a wide-spread Clojure library for monadic programming.
Similarly, we show that Typed Clojure helps verify code using Java interoperability is correct, and is
a useful tool to show misusing of Java's \emph{null}.

>>>>

Typed Clojure is intended to be of practical use to Clojure programmers.
We show that Typed Clojure helps verify the absence of errors in Clojure code written in sophisticated
programming styles by porting most of a wide-spread Clojure library for monadic programming.
Similarly, we show that Typed Clojure helps verify code using Java interoperability is correct, and is
a useful tool to show misuse of Java's \emph{null}.
